\documentclass[11pt]{article}

%use one of the following package accordingly

%\usepackage[brazil]{babel} % for portuguese
\usepackage[english]{babel} % for english
%\usepackage[spanish]{babel} % for spanish

\usepackage[latin1]{inputenc} % for accents in portuguese
%\usepackage[utf8]{inputenc} % for accents in portuguese using Unicode 
%%
%%
%% PLEASE DO NOT MAKE CHANGES TO THIS TEMPLATE 
%% THAT CAUSE CHANGES IN THE FORMAT OF THE TEXT
%%
%%
\usepackage[centertags]{amsmath}
\usepackage{indentfirst,amsfonts,amssymb,amsthm}
\usepackage{cite}
\usepackage{pdfsync}
\usepackage{url}
\usepackage[bottom=1.5cm,top=1.5cm,left=3cm,right=2cm]{geometry}
\date{}

\begin{document}

%********************************************************
\title{Confluence of an Explicit Substitutions Calculus Formalized}

\author{
  {\large Fl\'avio L. C. de Moura}\thanks{flaviomoura@unb.br},
    {\large Leandro O. Rezende}\thanks{L-ordo.ab.chao@hotmail.com}\\
  {\small Departamento de Ci\^encia da Computa\c{c}\~ao}\\
  {\small Universidade de Bras\'ilia, Bras\'ilia, Brazil} }

\maketitle

\begin{abstract}
%the abstract should contain a maximum of 600 words
  Rewriting theory is a well established model of computation
  equivalent to the Turing machines, and the best-known rewriting
  system is the $\lambda$-calculus, the theoretical foundation of the
  functional paradigm of programming. Confluence is an important
  property related to the determinism of the results given by a
  rewriting system. In this work, which is still in progress, we
  formalize the confluence property of an extension of the
  $\lambda$-calculus with explicit substitutions following the steps
  in \cite{DK08,kes09}. Confluence is obtained through the Z
  property~\cite{zproperty}, and the first challenge of this work was
  to prove that an abstract rewriting system, i.e. a binary relation
  over an arbitrary set, that satisfies the Z property is
  confluent. The difficulty relies on the precise structure of the
  nested induction that needs to be done on the reflexive transitive
  closure generated by the divergence in the definition of
  confluence. The formalization is done in the Coq proof
  assistant~\cite{CoqTeam}, a system based on a constructive
  higher-order logic with a well developed extraction
  mechanism~\cite{Let2008}.

  In the $\lambda$-calculus, terms that only differ by the name of its
  bound variables are considered equal. This notion is known as
  $\alpha$-equivalence, which is a costly computational equivalence
  relation. Alternatives to $\alpha$-equivalence include the so called
  De Bruijn indexes~\cite{dB72}, where variables are represented by
  natural numbers. In De Bruijn notation terms have a unique
  representation, and hence there is no need of
  $\alpha$-equivalence. Nevertheless, defining a reduction in De
  Bruijn notation requires a non-trivial algebra for referencing and
  updating indexes. The Locally Nameless Representation
  (LNR)~\cite{Ch11} is a framework that takes the advantages of the
  two notations: bound variables are represented as De Bruijn indexes,
  while free variables uses names. The original framework uses
  classical logic and was built for representing pure $\lambda$-terms,
  therefore we decided to take some of its constructions (which are
  not based on classical logic) and extend it with a new operator for
  the substitution operation in such a way that our framework is
  constructive. This is important because one of the goals of this
  work is the generation of certified code via the extraction
  mechanism of Coq.

  Our formalization is based on the paper~\cite{DK08}, where the
  $\lambda$ex-calculus is defined. Another challenging step of this
  formalization is that the $\lambda$ex-calculus defines an equational
  theory based on the fact that reduction is done modulo permutation
  of independent substitutions. In order to avoid an explicit
  manipulation of permutation of independent substitutions, we use the
  generalized rewriting facilities of
  Coq~\cite{Sozeau09}. Nevertheless, the generated equivalence
  relation needs to be defined over every expression defined by the
  LNR grammar, and not only over $\lambda$-terms with explicit
  substitutions. In order to circumvent this problem, we proved that
  the reduction relation defined by the calculus in LNR modulo
  permutations of independent substitutions is restricted to
  $\lambda$-terms with explicit substitutions.

  The formalization is available at
  \url{https://github.com/flaviodemoura/Zproperty.git} and is divided
  in two files:
  \begin{enumerate}
  \item The file {\tt ZtoConfl.v} contains the proof that an abstract
    rewriting system $R$ that satisfies the Z property is confluent;
  \item The file {\tt lex.v} contains the current status of the
    formalization showing that the calculus in LNR satisfies the
    Z-property, and hence is confluent.
  \end{enumerate}  
\end{abstract}

\bibliography{references}
 \bibliographystyle{plain}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
