\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,authorcolumns]{lipics-v2019}
\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}
%\usepackage[british]{babel}
%\usepackage{lmodern}
%\usepackage{fullpage}
\usepackage{coqdoc}
\usepackage{mathpartir}
% \usepackage{amsmath,amssymb}
% \usepackage{amsfonts, color}
% \usepackage{url}
% \usepackage{alltt}
\usepackage[all]{xy}
% \usepackage{hyperref}
% \usepackage{graphicx}
% \usepackage[backend=bibtex,style=numeric-comp,sorting=none,firstinits=true]{biblatex}
% \usepackage{tikz}
% \usetikzlibrary{positioning,shapes,fit,arrows}

% \definecolor{myblue}{RGB}{56,94,141}

\bibliographystyle{plainurl}


% \newcommand{\flavio}[1]{{\color{red} #1}}
% \newcommand{\cflavio}[2]{{\color{red} {#2}}}

\newcommand{\tto}{\twoheadrightarrow}


\title{Confluence via the Z Property in Coq}

\titlerunning{Confluence via the Z Property }

\author{Flávio L. C. de Moura\footnote{Corresponding author}}{Departamento de Ci\^encia da Computacao, Universidade de Brasilia \and
  \url{http://flaviomoura.mat.br}
}{flaviomoura@unb.br}{https://orcid.org/0000-0002-9390-5751}{}

\author{Leandro O. Rezende}{Departamento de }{L-ordo.ab.chao@hotmail.com}{}{}

\authorrunning{F. L. C. de Moura and L. O. Rezende}

\Copyright{Flávio L. C. de Moura and Leandro O. Rezende}

\ccsdesc[100]{Logic and verification}

\keywords{Abstract rewriting systems, Confluence, Z property, Iteractive theorem proving, Coq}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle
%\pagestyle{plain}

\begin{abstract}
  Rewriting theory is a well established model of computation
  equivalent to the Turing machines, and the most well known rewriting
  system is the $\lambda$-calculus. Confluence is an important and
  undecidable property related to the determinism of the computational
  process. Direct proofs of confluence are, in general, difficult to
  be done. Therefore, alternative characterizations of confluence can
  circumvent this difficulty for different contexts. This is the case
  of the so called Z property, which has been successfully used to
  prove confluence in several situations such as the
  $\lambda$-calculus with $\beta\eta$-reduction, extensions of the
  $\lambda$-calculus with explicit substitutions, the
  $\lambda\mu$-calculus, etc. In this work we present a constructive
  proof that the Z property implies confluence. The known proofs of
  this fact usually rely on the law of the excluded middle. In
  addition, we formalized our proof and an extension of the Z
  property, known as the Compositional Z, in the Coq proof assistant.
\end{abstract}

\section{Introduction}



Confluence is an important and undecidable property concerning the
determinism of the computational process. In this sense one says that
a program is confluent if every two ways of evaluating it result in
the very same answer. In the particular case of Abstract Rewriting
Systems (ARS), which are the focus of this work, confluence can be
beautifully expressed by diagrams as we will see in the next section.

The contributions of this work are as follows:
\begin{itemize}
\item We present a constructive proof of the fact that the Z property
  implies confluence. The result is a new proof based on nested
  induction that differs from the well known ones
  (\cite{kes09,ZPropertyDraft}) in what it does not rely on the law of
  the excluded middle.
\item The proof that the Z property implies confluence is formalized
  in the Coq proof assistant, and the presentation is made
  interleaving English followed by the corresponding Coq code. In this
  way, the annotations are done directly in the Coq files using the
  coqdoc annotation style. We believe that this approach is
  interesting for those that are not familiar to the Coq proof
  assistant because the English explanation followed by the
  corresponding Coq code gives a good idea on how they relate to each
  other. This discipline also induces a better organization of the
  formalization.
\item We formalize an extension of the Z property, known as
  compositional Z property, as presented in
  \cite{Nakazawa-Fujita2016}.
\end{itemize}

\input{../src/ZtoConfl.tex}

\section{Conclusion}



\bibliography{/Users/flaviomoura/idrive/bibliography/references.bib}

\end{document}
